### 2016-09-15 paper/3_gilbert_features.tex
Refactored paper structure

%!TEX root=main.tex
\subsection{Gilbert Typing}
\label{cha:gilberttyping}

@@

>>>>

\section{Gilbert Language}
\label{cha:gilbertlanguage}

Prior to the actual start of development of Gilbert, we had to decide which higher-level language Gilbert should support.
Since Gilbert aims to be a sparse linear algebra environment, R and MATLAB are the natural candidates.
In fact, R and MATLAB are quite similar and are both widely used in academics as well as industry to develop mathematical programs.
Due to their expressiveness and their rich library support with myriads of mathematical functions, they are just as well suited for quick prototyping as it is for full-fledged development.
In combination with their ease of use, they quickly became the standard in industry for numerical processing.
Our final choice fell on MATLAB, because of our higher familiarity with this language.
Yet, we believe that there are no fundamental reasons which would prohibit the support of R.
In fact, the translation from one language to the other should be straight-forward on the compiler level.

Another reason to imitate an existing language and not to devise a new language geared towards parallel execution is the laziness of people.
Since its part of human nature to be sluggish, people are initially unwilling to learn new things or to re-adapt.
Therefore, Gilbert was conceived to require as little re-adaption as possible of users familiar with MATLAB.
We believe that this aspect is a crucial property of Gilbert in order to be successfully adopted by the statistical community.
Furthermore, a neat side effect is that existing MATLAB code can almost seamlessly be ported to Gilbert and thus benefits instantaneously from the computational power of a large shared nothing cluster.
This feature becomes particularly relevant considering the huge existing code base.
Re-coding parts of it would be prohibitive, because of the required labor and provoked costs.

In the following sections, we will describe the key aspects of the Gilbert language and explain the design decisions taken.
Subsequently, we will give a formal specification of the Gilbert language and describe the used parser.

\subsection{Language Features}
\label{sec:languageFeatures}

In order to not be overwhelmed by the complexity of MATLAB but still support a full functional subset of it, Gilbert was restricted to support the following language features and data types.
We decided to keep the feature set as small as possible to concentrate on the parallelization instead of supporting hardly used language features.
The elementary data type of linear algebra and therefore also of MATLAB is a matrix.
Gilbert supports arbitrary $2$-dimensional matrices whose elements can be \emph{double} or \emph{boolean}.
Vectors are not represented by a special type but instead are treated as a matrix.
Additionally, scalar \emph{double} and \emph{boolean} values are supported.
The reason to introduce a \emph{boolean} type is the control flow of loops, being covered in a later paragraph.
Since Gilbert has to interact with data stored on disk, strings are supported.

Gilbert also implements cell array types.
A cell array consists of indexed data containers, called cells.
Each cell can contain an individual data type.
Therefore, cell arrays can be used to pass multiple data items combined as one argument to a function or to obtain multiple items from a function.
The latter aspect is particularly important, because Gilbert only supports functions with a single return value.
A cell array is defined using curly braces and commas to separate individual values.
The cells can be accessed by an index appended in curly braces to the cell array variable.
\Cref{lst:cellArray} shows how to define and access them.
The formal definition of the syntax of cell arrays can be found in \cref{sec:languageGrammar}.
In contrast to MATLAB's cell arrays, though, Gilbert only allows to have $1$-dimensional arrays.
However, this constraint does not impose a serious restriction, since all multi-dimensional arrays can be transformed into $1$-dimensional arrays.

\begin{listing}[!h]
  \begin{CenteredBox}
    \begin{lstlisting}[language=Matlab,
        commentstyle=\color{black},
        stringstyle=\color{black},
    ]
c = {true, 2*2, 'cell', 'array'};
b = c{1} & false; % = false
d = c{2} ^ 2; % = 16
s = {c{4}, c{3}}; % = {'array', 'cell'} 
    \end{lstlisting}
  \end{CenteredBox}
  \caption{Cell array usage in Gilbert. Definition of a 4 element cell array which is accessed subsequently.}
  \label{lst:cellArray}
\end{listing}

Gilbert supports the basic linear algebra operations defined on matrices and scalars.
They include among others the common operations \code{+}, \code{-}, \code{/} and \code{*}, whereas \code{*} denotes the matrix-matrix multiplication and all other operations are interpreted cellwisely.
The cellwise multiplication is indicated by a point preceding the operator as it is common for cellwise operations in MATLAB.
Gilbert also supports comparisons operators such as \code{>}, \code{>=}, \code{==} and \code{\textasciitilde=}.
These operators become handy for realizing dynamic termination criteria for loops.
The full set of supported mathematical operations can be found in appendix 1.

Besides the basic arithmetic operations, the user can also define named functions and anonymous functions.
This feature not only enables a better structure of the code but also allows to express second-order functions.
The syntax of anonymous functions adheres to the MATLAB syntax.
An anonymous function which calculates the sum of its input squares could be defined as follows: \code{@(x,y) x*x + y*y}.
A formal definition can be found in \cref{sec:languageGrammar}.

An important aspect of MATLAB are loops.
MATLAB permits the user to express \code{for} and \code{while} loops.
However, these loops are quite powerful in the sense that they allow iterations with side effects.
The problem of parallelization of iterations with side effects is that the referenced external state has to be maintained.
This circumstance makes preprocessing and execution unnecessarily complex.
For the sake of simplicity, Gilbert is limited to a different loop mechanism.
Gilbert offers a fixpoint operator \code{fixpoint}, which iteratively applies a given update function $f$ on the previous result of $f$, starting with an initial value $x$ at iteration $0$.
Thus, the $n^{th}$ iteration is equivalent to applying the function $f$ $n$ times to $x$: 
\begin{displaymath}
  n^{th}\text{ iteration}\equiv\underbrace{f(f(\ldots(f(x))\ldots))}_{\text{$n$ times}}
\end{displaymath}

In order to terminate the fixpoint operation, the operator provides two mechanisms.
First of all, the user has to specify a maximum number \code{m} of iterations after which the loop is terminated.
This mechanism is henceforth denoted as the \emph{static termination criterion}.
Such a termination criterion is often not sufficient for machine learning algorithms, because they usually continue their computations until they have reached a certain convergence criterion.
Since the number of iterations necessary to reach this state is not always known a priori, it is called the \emph{dynamic termination criterion}.
Gilbert offers support for the dynamic termination criterion.
The user can provide a convergence function \code{c} to the fixpoint operator.
The convergence function is called with the previous and current fixpoint value and returns a boolean value, indicating whether the termination criterion has been fulfilled or not.
Thus, the fixpoint operator terminates either if convergence was detected or if the maximum number of iterations is exceeded.
Consequently, the fixpoint operator is defined as follows:
\begin{equation}
fixpoint: \underbrace{T}_{\text{\code{x}}} \times \left( \underbrace{T \rightarrow T}_{\text{\code{f}}} \right) \times \underbrace{\mathbb{N}}_{\text{\code{m}}} \times \left(\underbrace{T\times T \rightarrow \mathbb{B}}_{\text{\code{c}}} \right) \rightarrow T
\label{eqn:fixpoint}
\end{equation}
with $T$ being a generic type variable.

In fact, the fixpoint operator replaces iterations by recursions whereas the update function $f$ is pure.
At this point Gilbert breaks with existing MATLAB code.
To make Gilbert a real subset of MATLAB, the fixpoint operator would have to be integrated into MATLAB.
The integration could easily be done by providing an external library with the definition of the fixpoint function.

Even though the iteration operator restricts the set of valid programs in Gilbert, it is still expressive enough to support a wide variety of programs.
Moreover, all MATLAB programs can be transformed so that the for and while loops are replaced by the fixpoint operator.
This replacement is simply achieved by passing all data that is read or written to as parameters to the update function.
The update function returns the same set of variables, just with updated values.
Such an exemplified transformation can be seen in \cref{fig:for2Fixpoint}.
Here we can see that all data that is operated on is passed to the fixpoint operator as a cell array.
Lines $1-2$ of \cref{fig:for2Fixpoint:fixpoint} define the update function \code{f}.
The parameter \code{x} is a cell array whose first entry contains the accumulator \code{A} and second entry is the loop counter \code{i}.
In line $2$ we see the returned cell array value of the anonymous function.
The first entry is the sum of the current accumulator value and the loop counter.
The second entry is the incremented loop counter for the next iteration.
Line $3$ calls the fixpoint operator with the initial cell array value, the update function and the maximum number of iterations.
The final result is retrieved from the final cell array value in line $4$.

\begin{listing}
  \centering
  \begin{sublisting}{.4\linewidth}
    \begin{lstlisting}[language=Matlab,
      commentstyle=\color{black},
      stringstyle=\color{black},
      keywordstyle=\color{black}\bfseries,
    ]
A = 0;
for i = 1:10
  A = A + i;
end
    \end{lstlisting}
    \caption{For loop}
    \label[listing]{fig:for2Fixpoint:for}
  \end{sublisting}
  \begin{sublisting}{.5\linewidth}
      \begin{lstlisting}[language=Matlab]
f = @(x) ...
  {x{1} + x{2}, x{2} + 1};
r = fixpoint({0,1}, f, 10);
A = r{1};
    \end{lstlisting}
    \caption{Fixpoint}
    \label[listing]{fig:for2Fixpoint:fixpoint}
  \end{sublisting}
  \caption{Transformation from Matlab for loop \subref{fig:for2Fixpoint:for} to Gilbert fixpoint \subref{fig:for2Fixpoint:fixpoint} formulation. Essentially, all iteration data is combined and passed as a cell array value to the update function.}
  \label{fig:for2Fixpoint}
\end{listing}

\subsection{Language Grammar}
\label{sec:languageGrammar}

We could not find an official specification of the MATLAB language.
Therefore, we defined our own grammar with the goal to imitate MATLAB as closely as possible.
Even though we extensively tested the conformity of the language, we cannot guarantee that it is completely equivalent.
In the following, we give the Backus-Naur form (BNF) of the Gilbert language.

%\setlength{\grammarindent}{12em} % increase separation between LHS/RHS
\begin{lstlisting}
<program> ::= <stmtOrFuncList>

<stmtOrFuncList> ::= (<stmt> | <funcDef>)*

<stmt> ::= <assignment> | <expression>

<assignment> ::= <lhs> '=' <rhs>

<lhs> ::= <identifier>

<rhs> ::= <expression>

<funcDef> ::= 'function' <funcValues>? <identifier> <funcParams> <funcBody> 'end'

<funcValues> ::= <identifier> '='
\alt '[' <identifier> (',' <identifier>)* ']' '='

<funcParams> ::= '(' ')'
\alt '(' <identifier> (',' <identifier>)* ')'

<funcBody> ::= <stmtOrFuncList>

<expression> ::= <aexp1>

<aexp1> ::= <aexp2> ('||' <aexp2>)*

<aexp2> ::= <aexp3> ('\&\&' <aexp3>)*

<aexp3> ::= <aexp4> ('|' <aexp4>)*

<aexp4> ::= <aexp5> ('\&' <aexp5>)*

<aexp5> ::= <aexp6> (<compOp> <aexp6>)*

<aexp6> ::= <aexp7> (':' <aexp7>)*

<aexp7> ::= <aexp8> (<addOp> <aexp8>)*

<aexp8> ::= <aexp9> (<multOp> <aexp9>)*

<aexp9> ::= <prefixOp> <aexp9> | <aexp10>

<aexp10> ::= <aexp11> (<expOp> <aexp11>)*

<aexp11> ::= <unaryExpression> <postfixOp>?

<unaryExpression> ::= <elemExpression> | '(' <expression> ')'

<elemExpression> ::= <funcApplication> \alt <cellExpression> \alt <identifier> \alt <scalar> \alt <booleanLiteral> \alt <matrix> \alt <stringLiteral> \alt <anonymousFunc> \alt <funcReference>

<funcApplication> ::= <identifier> '(' ')'
\alt <identifier> '(' <expression> (',' <expression>)* ')'

<funcReference> ::= '@' <identifier>

<cellExpression> ::= <cellArray> | <cellArrayIndexing>

<cellArray> ::= '\{' <expression> (',' <expression>)* '\}'

<cellArrayIndexing> ::= <identifier> ('\{' <numericLiteral> '\}')+

<scalar> ::= <numericLiteral>

<matrix> ::= '[' <matrixRow> (<newlineOrSemicolon> <matrixRow>)* ']'

<matrixRow> ::= <expression> (',' <expression>)*

<prefixOp> ::= '+' | '-'

<multOp> ::= '*' | '/' | '.*' | './'

<addOp> ::= '+' | '-'

<compOp> ::= "'<'" | "'<='" | "'>'" | "'>='" | "'=='" | "'~='"

<postfixOp> ::= '.'' | '''

<expOp> ::= '.\textasciicircum'

<newlineOrSemicolon> ::= NL | ';'

\end{lstlisting}

For the sake of simplicity, we have left out most of the whitespace and line break handling.
Furthermore, we used notations known from regular expressions to shorten the grammar specification.
We use parentheses to group subexpressions and apply the following multiplicity specifiers: \code{*}, \code{+} and \code{?}.
The semantics of these specifiers can be found in \cref{tab:multiplicity}.

\begin{table}[!h]
  \centering
  \begin{tabular}{c|l}
  Multiplicity specifier & Meaning\\
  \hline
  $\langle exp \rangle$* & $\langle exp\rangle$ can appear multiple times \\
  $\langle exp\rangle$+ & $\langle exp\rangle$ can appear multiple times but at least once \\
  $\langle exp\rangle$? & $\langle exp\rangle$ can appear exactly once or not at all
  \end{tabular}
  \caption{Multiplicity specifier used to specify the Gilbert language. They are borrowed from regular expressions.}
  \label{tab:multiplicity}
\end{table}

The different \emph{aexp} non-terminals are used to model the precedence order of the mathematical operators.
An overview of the precedence can be found in \cref{tab:precedences}.
The $'$ operator denotes the transpose of the operand.

\begin{table}[!h]
  \centering
  \begin{tabular}{l|l}
  Operator class & Operators\\
  \hline
  Short circuit logical or & $||$\\
  Short circuit logical and & $\&\&$\\
  Logical or & $|$ \\
  Logical and & $\&$ \\
  Comparators & $>$, $>=$, $<$, $<=$, $==$, \textasciitilde$=$\\
  Range operator & $:$ \\
  Addition and Subtraction & $+$, $-$\\
  Multiplication and Division & $*$, $/$\\
  Prefix operator & $+$, $-$\\
  Exponentiation & \textasciicircum\\
  Postfix operator & $'$
  \end{tabular}
  \caption{Precedence order of mathematical operators in ascending order.}
  \label{tab:precedences}
\end{table}

The non-terminals \emph{identifier}, \emph{numericLiteral}, \emph{stringLiteral} and \emph{booleanLiteral} are tokens generated by the lexer. The corresponding tokens are defined by regular expressions and can be found in \cref{tab:tokens}.

\begin{table}[!h]
  \centering
  \begin{tabular}{l|l}
  Token name & Regular expression\\
  \hline
  identifier & [a-zA-Z] [a-zA-Z0-9\_]*\\
  numericLiteral & [+-]? [0-9]+ (.[0-9]*)? ([eE][+-]?[0-9]+)? | [+-]? .[0-9]+ ([eE][+-]?[0-9]+)? \\
  stringLiteral & "[\textasciicircum"]*" | '[\textasciicircum']*'\\
  booleanLiteral & true | false
  \end{tabular}
  \caption{Definition of generated tokens by regular expressions.}
  \label{tab:tokens}
\end{table}

\subsection{Parser}

In order to implement Gilbert, a parser had to be selected which is powerful enough to parse the Gilbert language.
Luckily, the language is rather simple and since none of the production rules is left recursive, we can use a simple LL(n) parser.
This aspect becomes important when we justify our implementation choices in \cref{cha:implementation}.

The attentive reader will have noticed that our language specification contains an ambiguity.
Consider, for example, the following valid MATLAB code \code{A' + B'} where \code{A} and \code{B} are matrices.
The code should produce the sum of two transposed matrices.
However, on the lexer level the two transpose operators will be recognized as the enclosing apostrophes of a string literal.
Therefore, the lexer will produce an identifier token followed by a string literal token.
Eventually, these tokens will cause a parser error, because there is no production rule consuming such a combination of tokens.

In order to solve this problem, we have to differentiate a transpose operator from an apostrophe belonging to a string literal.
Since the transpose operator always follows after an expression, we can check that the previously detected token is an identifier, closing parentheses, closing bracket or closing brace.
If one of these cases is true, then the apostrophe is considered to be a transpose operator.
Otherwise, it belongs to a string literal.

\subsection{Gilbert Typing}
\label{cha:gilberttyping}

@@

\section{Intermediate Representation}
\label{cha:intermediaterepresentation}

After parsing and typing of the source code is done, we have all the information needed to execute the program in a distributed fashion.
However, we do not compile the code directly to the respective format to let it run on a parallel data processing system.
Instead, we compile it to an intermediate representation.
Even though, the intermediate code adds an additional representation and thus more complexity to maintain, the advantages clearly outweigh the disadvantages.
Moreover, this approach is similar to the approach the JVM and the .Net frameworks pursue.
The Java and .Net code are compiled to a format which can be executed platform independently.

The additional abstraction layer allows Gilbert to apply language independent optimization in a coherent manner and independently of the actually used front end language.
This implementation aspect gives the flexibility to easily extend Gilbert to support other linear algebra languages as well.
The next natural candidate for a supported front end language would surely be R.
Moreover, a higher-level abstraction of the mathematical operations holds more potential for algebraic optimization.
One class of important transformations is the reordering of operations and the determination of the execution order.
Often, they are based on the commutative and associative property of the corresponding operations.
An obvious optimization would be the execution order of multiple matrix multiplications as indicated in \cref{sec:MatrixDimensionInference}.
Furthermore, the propagation and elimination of transpose operations offers further optimization potential.

The intermediate representation is designed with two goals in mind:
First of all, it has to be expressive enough to represent in a general way a wide variety of MATLAB programs.
Therefore, the intermediate format has to include the operational primitives of linear algebra as well as an iteration abstraction.
With these building blocks at hand, it is already possible to express a multitude of iterative machine learning algorithms.
As pointed out in \cref{sec:languageFeatures}, Gilbert does not provide a direct \code{for} or \code{while} loop, but instead the \code{fixpoint} operator.
The second design goal was simplicity.
It is of particular interest to keep the set of intermediate operators as small as possible, because it would alleviate a possible optimization step prior to execution.

\subsection{Specification}

The intermediate format consists of a set of operators to represent the different linear algebra operations.
Every operator has a distinct result type and a set of parameters which are required as inputs.
\Cref{fig:depTreeMMIdOnes} shows how a matrix multiplication between an identity matrix $Id \in \mathbb{R}^{10\times10}$ and a matrix $Ones \in \mathbb{R}^{10\times10}$ filled with $1$s would be represented.
On the level of the intermediate format we distinguish between the following types: \code{MatrixType}, \code{ScalarType}, \code{CellArrayType}, \code{StringType} and \code{FunctionType}.

\begin{figure}
  \begin{center}
    \Tree [.MatrixMult [.eye [.scalar 10 ] [.scalar 10 ] ] [.ones [.scalar 10 ] [.scalar 10 ] ] ]
  \end{center}
  \caption{Dependency tree of a matrix multiplication between an identity matrix and a matrix filled with 1s.}
  \label{fig:depTreeMMIdOnes}
\end{figure}

The matrix type is the elementary type and represents all sorts of matrices.
A matrix contains information about its size, namely the number of rows and columns, and about its element type.
The element type can be any supported type but currently only operations for double or boolean matrices are implemented.

The scalar type is a superset containing the \code{DoubleType} and \code{BooleanType}.
Usually, the double type is used for all linear algebra operations.
Boolean scalars are mainly used for convergence criteria of iterations.

The cell array type contains a list of types specifying the types of its cell entries.
Cell arrays are usually used to return multiple values from a function.
Thus, they can conceptually be seen as tuples.
In contrast to the MATLAB implementation, Gilbert only supports $1$-dimensional cell arrays.

The string type was included to be able to load matrices from HDFS or local disk.
In order to load a matrix, the user has to define the location of the data, given as a string.
Besides this usage, strings are usually not necessary for linear algebra operations.

Last but not least, Gilbert also supports function types.
Function types are necessary to implement second-order functions such as the fixpoint operator.
The user can pass defined functions around like any other value.
Thus, functions are considered first-class citizens in Gilbert.

The set of intermediate operators can be distinguished into three categories: \emph{Creation operators}, \emph{transformation operators} and \emph{Output operators}.
All of these categories will be explained in the following subsections.

\subsubsection{Creation Operators}

Creation operators generate or load some data depending on the used function.
A list of all supported operators with its type definition and a short explanation can be found in \cref{tab:creatingOperators}.
Each type definition consists of the input types and the result type of an operator:
\begin{displaymath}
  \underbrace{type1 \times \ldots \times typeN}_{\text{Types of input parameters}} \rightarrow \underbrace{resultType}_{\text{Type of result value}}
\end{displaymath}
Even though the input parameters for the matrix dimensions are denoted by doubles, the user should provide integer values.
Internally, the double parameters are converted to integers.
The type in brackets after the matrix type indicates the element type.

The \code{load} operator takes a path to a stored matrix on disk and the corresponding number of rows and columns as input parameters.
It then reads the stored data and loads the matrix into the program context.
The \code{eye} operator, known from Matlab, generates an identity matrix of the requested size, number of rows and columns.
The operator also supports non quadratic result shapes.
Another well-known operator is \code{zeros}.
It generates a matrix of the given size which is initialized with zeros.
And the last creation operator is \code{randn} which generates a random matrix.
\code{Randn} takes the number of rows and columns of the resulting matrix and the mean and the standard deviation of a Gauss distribution.
The specified Gauss distribution is used to generate random values for the resulting matrix.

\begin{table}
  \centering
  \begin{tabular}{l|l|l}
  Operator & Type & Explanation\\
  \hline
  \code{load} & $string \times double \times double \rightarrow matrix[double]$ & Loading a matrix from disk\\
  \code{eye} & $double \times double \rightarrow matrix[double]$& Creating an identity matrix\\
  \code{zeros} & $double \times double \rightarrow matrix[double]$ & Creating a zero matrix\\
  \code{randn} & $double \times double \rightarrow matrix[double]$ & Creating a random matrix
  \end{tabular}
  \caption{Creating operators of Gilbert, their types and a short explanation.}
  \label{tab:creatingOperators}
\end{table}

\subsubsection{Transformation Operators}

The transformation operators constitute the main abstraction of the linear algebra operations.
They group operations with similar properties and thus allow an easier reasoning and optimization of the underlying program.
The list of all transformation operators and their types can be found in \cref{tab:transformingOperators}.

\begin{table}
  \centering
  \begin{tabular}{l|l}
    Operator & Type\\
    \hline
    \code{UnaryScalarTransformation} & $scalar \times unaryScalarOp \rightarrow scalar$\\
    \code{ScalarScalarTransformation} & $scalar \times scalar \times scalarOp \rightarrow scalar$\\
    \code{Transpose} & $matrix \rightarrow matrix$\\
    \code{ScalarMatrixTransformation} & $scalar \times matrix \times smOp \rightarrow matrix$ \\
    \code{MatrixScalarTransformation} & $matrix \times scalar \times smOp \rightarrow matrix$ \\
    \code{CellwiseMatrixTransformation} & $matrix \times unaryScalarOp \rightarrow matrix$ \\
    \code{CellwiseMatrixMatrixTransformation}& $matrix \times matrix \times scalarOp \rightarrow matrix$ \\
    \code{MatrixMult} & $matrix \times matrix \rightarrow matrix$\\
    \code{VectorwiseMatrixTransformation} & $matrix \times vectorwiseOp \rightarrow matrix$ \\
    \code{AggregateMatrixTransformation} & $matrix \times aggregateOp \rightarrow scalar$ \\
    \code{FixpointIterationMatrix} & $matrix \times \left( matrix \rightarrow matrix \right) \times double \times$ \\
    &$\quad \left( matrix \times matrix \rightarrow boolean \right) \rightarrow matrix$ \\
    \code{FixpointIterationCellArray} & $cellarray \times \left( cellarray \rightarrow cellarray \right) \times double$ \\
    &$\quad \times \left(cellarray \times cellarray \rightarrow boolean\right)$\\
    &$\quad \rightarrow cellarray$
  \end{tabular}
  \caption{Transforming operators of Gilbert and their types.}
  \label{tab:transformingOperators}
\end{table}

The \code{UnaryScalarTransformation} takes a single scalar value and applies an unary operation on it.
The list of all supported operations can be found in \cref{tab:supportedOperations}.
The operation $binarize$ has the following semantics:
\begin{displaymath}
  binarize(x) : x \mapsto 
  \begin{cases}
    0 &\text{if } x = 0\\
    1 &\text{else}
  \end{cases}
\end{displaymath}

The applicable operations to the \code{ScalarScalarTransformation} depend on the input parameter types.
The operators $+,-,*,/,pow(\cdot, \cdot)$ take two double values and produce a new double value.
The comparison operators, on the contrary, produce a boolean result value.
The logical operations require two boolean values and produce a boolean value as the result.
Gilbert supports short-circuit logical operators $\&\&$ and $||$ as well as $\&$ and $|$ evaluating all of their inputs prior to computing the result.

The \code{ScalarMatrixTransformation} and \code{MatrixScalarTransformation} have an almost identical support of operations.
The only difference is that the former transformation does not allow to calculate the power of a scalar where the exponent is a matrix.

The \code{VectorwiseMatrixTransformation} applies an operation on each row vector of the given matrix.
A vectorwise operation produces a scalar value for each row vector.
Thus, the resulting type of such an operation is an one-column matrix.
Gilbert can currently retrieve the maximum and minimum of each row vector.
Furthermore, one can calculate the $2$-norm of each row vector:
\begin{displaymath}
  \norm{\left(a_{i1}, \ldots, a_{in}\right)^T}_2 = \sqrt{\sum_{j=1}^{n} a_{ij}^2}
\end{displaymath}

Similar to the vectorwise operations, the \code{AggregateMatrixTransformation} applies an operation to all matrix entries producing a single scalar result value.
The aggregation operation can compute the maximum, minimum and the sum of all matrix entries.
Furthermore, it supports the Frobenius norm:
\begin{displaymath}
  \norm{A}_2 = \sqrt{\sum_{i=1}^{m}\sum_{j=1}^{n} a_{ij}^2}
\end{displaymath}

\begin{table}
  \centering
  \begin{tabular}{l|l}
  Operation type & Operations\\
  \hline
  $unaryScalarOp$ & $-$, $|\cdot|$, $binarize(\cdot)$\\
  $scalarOp:double \times double \rightarrow double$ & $+,-,*,/,pow(\cdot, \cdot)$\\
  $scalarOp:double \times double \rightarrow boolean$ & $>,>=,<,<=,==,$\textasciitilde$=$\\
  $scalarOp:boolean \times boolean \rightarrow boolean$ & $\&, |, \&\&, ||$\\
  $smOp:matrix[double] \times double \rightarrow matrix[double]$ & $+,-,*,/,pow(\cdot, \cdot)$\\
  $smOp:matrix[double] \times double \rightarrow matrix[boolean]$ & $>,>=,<,<=,==,$\textasciitilde$=$\\
  $smOp:matrix[boolean] \times boolean \rightarrow matrix[boolean]$ & $\&, |, \&\&, ||$\\
  $smOp:double \times matrix[double] \rightarrow matrix[double]$ & $+,-,*,/$\\
  $smOp:double \times matrix[double] \rightarrow matrix[boolean]$ & $>,>=,<,<=,==,$\textasciitilde$=$\\
  $smOp:boolean \times matrix[boolean] \rightarrow matrix[boolean]$ & $\&, |, \&\&, ||$\\
  $vectorwiseOp$ & $min, max, norm2$\\
  $aggregateOp$ & $min, max, sumAll, norm2$
  \end{tabular}
  \caption{Supported transformation operations.}
  \label{tab:supportedOperations}
\end{table}

The iteration mechanism is represented by the \code{FixpointIterationMatrix} and \\\code{FixpointIterationCellArray} operators.
The former operator is used for a fixpoint operation on matrices and the latter on cell arrays.
Both operators follow the semantics and have the same parameters as the fixpoint abstraction presented in \cref{eqn:fixpoint}.
A fixpoint operator receives an initial value, an update function, the maximum number of iterations and a convergence function.
The update function is called with the current value and returns the value for the next iteration.
The convergence function is called with the current and the previous value.
Based on these values it can check for convergence.
For further details see \cref{sec:languageFeatures}.

\subsubsection{Writing Operators}

The writing operators are used to make the computed results accessible to the user by writing them back to disk.
There exists a writing operation for each supported type.
The list of all writing operations can be seen in \cref{tab:writingOperations}.

\begin{table}[!h]
  \centering
  \begin{tabular}{l|l}
    Operator& Explanation\\
    \hline
    \code{WriteMatrix} & Writes the matrix to disk\\
    \code{WriteScalar} & Writes the scalar value to disk\\
    \code{WriteString} & Writes the string value to disk\\
    \code{WriteCellArray} & Writes the contents of the cell array to disk \\
    \code{WriteFunction} & Writes the intermediate execution plan to disk
  \end{tabular}
  \caption{Supported writing operators.}
  \label{tab:writingOperations}
\end{table}

\subsubsection{Compilation Example}

The intermediate representation is the target format of the compilation process after it has been parsed and typed.
In order to illustrate the compilation process and to get a feeling for the intermediate representation, we will compile an example program.
For this purpose, we have chosen the well-known PageRank algorithm~\cite{page:1999a}.
The Matlab implementation can be seen in \cref{lst:PageRank}.

\begin{listing}[h!]
  \begin{CenteredBox}
    \begin{lstlisting}[language=Matlab,
    commentstyle=\color{black},
      stringstyle=\color{black},
      keywordstyle=\color{black}\bfseries,
      morekeywords={ones, fixpoint},
      deletekeywords={eps},
      ]
numVertices = 10;
% load network matrix
N = load("network.csv", numVertices, numVertices);
% create the adjacency matrix
A = spones(N);
% outdegree per vertex
d = sum(A, 2);
% create the column-stochastic transition matrix
T = (diag(1 ./ d) * A)';
% initialize the ranks
r_0 = ones(numVertices, 1) / numVertices;
e = ones(numVertices, 1) / numVertices;
% update function
f = @(r) (.85 * T * r + .15 * e)
eps = 0.1;
c =  @(prev, cur) norm(prev-cur,2) <= eps
% PageRank calculation
fixpoint(r_0, f, 20, c);
    \end{lstlisting}
  \end{CenteredBox}
  \caption{Matlab PageRank implementation.}
  \label{lst:PageRank}
\end{listing}

We start at the fixpoint operation in line $20$ and build our intermediate representation bottom up.
Since the fixpoint operation is given an initial matrix, it will be compiled to a \\\code{FixpointIterationMatrix} operator.
The intermediate code can be seen in \cref{fig:irFixpoint}.

\begin{figure}[!h]
  \centering
  \Tree [.FixpointIterationMatrix r\_0 f [.scalar 20 ] c ]
  \caption{Intermediate representation of PageRank's fixpoint operation.}
  \label{fig:irFixpoint}
\end{figure}

The translation from line $20$ to the intermediate code is straight forward.
The actual compilation process would replace the variables \code{r\_0}, \code{f} and \code{c} with their actual definition.
However, for the sake of simplicity, we have left them in place and will gradually refine them.

The variable \code{r\_0} denotes the initial column vector of the PageRank iteration.
Its resulting code from the compilation process can be seen in \cref{fig:irR0}.

\begin{figure}[!h]
  \centering
  \Tree [.MatrixScalarOperation [.ones [.scalar 10 ] [.scalar 1 ] ] [.scalar 10 ] '/' ]
  \caption{Intermediate representation of PageRank's \code{r\_0}.}
  \label{fig:irR0}
\end{figure}

The update function \code{f} is defined in line $15$ as an anonymous function.
In order to represent functions in the intermediate format, we have to introduce a new operator \code{function} which takes the number of parameters and the body of the function.
Additionally, all parameters of the function are represented by special intermediate code operators: \code{MatrixParameter}, \code{ScalarParameter}, \code{StringParameter} and \code{FunctionParameter}.
Every parameter operator gets an index assigned which identifies the corresponding argument with which it has to be replaced upon instantiation of the function.
The compiled code of \code{f} can be found in \cref{fig:irF}.

\begin{figure}[!h]
  \centering
  \Tree [.function 1 [.CellwiseMatrixMatrixTransformation [.ScalarMatrixTransformation [.scalar .85 ] [.MatrixMult T [.MatrixParameter 0 ] ] '*' ] [.ScalarMatrixTransformation [.scalar .15 ] e '*' ] '+' ] ]
  \caption{Intermediate representation of PageRank's \code{f}.}
  \label{fig:irF}
\end{figure}

The matrix \code{T} is the transition matrix $T=(t)_{ij} \in \mathbb{R}^{10 \times 10}$.
The entry $t_{ij}$ indicates the probability of going from site $j$ to site $i$ if the user is currently on site $j$.
The dependency tree of the intermediate code of \code{T} is shown in \cref{fig:irT}.

\begin{figure}[!h]
  \centering
  \Tree [.Transpose [.MatrixMult [.diag [.ScalarMatrixTransformation [.scalar 1 ] [.sum A [.scalar 2 ] ] './' ] ] A ] ]
  \caption{Intermediate representation of PageRank's \code{T}.}
  \label{fig:irT}
\end{figure}

Here \code{diag} denotes a special operator which has a different behavior depending on the given argument.
If one provides a one-column matrix $d$ to \code{diag}, then a zero matrix with $d$ on its diagonal is created.
If one provides a matrix $m$ to \code{diag}, then the diagonal of $m$ is returned.
The operator \code{sum} takes a matrix and a dimension and computes the sums along the specified dimension.
Thus, the row sums will be computed in this case.

What is left to be compiled for \code{f} is the matrix \code{A}.
The compiled code can be seen in \cref{fig:irA}.

\begin{figure}[!h]
  \centering
  \Tree [.CellwiseMatrixTransformation [.load 'network.csv' [.scalar 10 ] [.scalar 10 ] ] 'binarize' ]
  \caption{Intermediate representation of PageRank's \code{A}.}
  \label{fig:irA}
\end{figure}

The Matlab function \code{spones}, which replaces nonzero sparse elements with ones while preserving the sparsity structure, is compiled into a CellwiseMatrixTransformation.

The \code{e} vector has the same representation as the initial PageRank vector \code{r\_0}.

\begin{figure}[!h]
  \centering
  \Tree [.MatrixScalarOperation [.ones [.scalar 10 ] [.scalar 1 ] ] [.scalar 10 ] '/' ]
  \caption{Intermediate representation of PageRank's \code{e}.}
  \label{fig:irE}
\end{figure}

The convergence function \code{c} is an anonymous function taking two parameters.
Its compiled version is shown in \cref{fig:irC}

\begin{figure}[!h]
  \centering
  \Tree [.function [.scalar 2 ] [.ScalarScalarTransformation [.norm [.CellwiseMatrixMatrixTransformation [.MatrixParameter 0 ] [.MatrixParameter 1 ] '-' ] [.scalar 2 ] ] [.scalar 0.1 ] '>=' ] ]
  \caption{Intermediate representation of PageRank's \code{c}.}
  \label{fig:irC}
\end{figure}

The \code{norm} operator, taking as first parameter the matrix $m$ and as second the norm parameter $p$, calculates the $p$-norm of $m$ over all matrix elements.
Internally this operator is compiled to the intermediate code shown in \cref{fig:irNorm}.

\begin{figure}[t]
  \centering
  \Tree [.function [.scalar 2 ] [.ScalarScalarTrans [.AggregateMatrixTrans [.CellwiseMatrixTrans [.MatrixScalarTrans [.MatrixParam 0 ] [.ScalarParam 1 ] 'pow' ] 'abs' ] 'sumAll' ] [.ScalarScalarTrans [.scalar 1 ] .ScalarParam 1 '/' ] 'pow' ] ]
  \caption{Intermediate representation of \code{norm}.}
  \label{fig:irNorm}
\end{figure}
